---
title: "p8105_hw3_xm2356"
author: "Xinyin Miao (xm2356)"
date: "2025-10-06"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6 )
```

```{r}
library(tidyverse)
```

# Problem 1
```{r}
library(p8105.datasets)
data("instacart")
```

The instacart dataset is a large dataframe containing `r nrow(instacart)` observations and `r ncol(instacart)` variables. Each observation represents a single item within a customer's order. Key variables include:

order_id: A unique identifier for each order.
product_id: A unique identifier for each product.
product_name: The name of the product.
aisle_id and aisle: The aisle number and name where the product is located.
department_id and department: The department number and name.
order_dow: The day of the week the order was placed (0 = Sunday).
order_hour_of_day: The hour of the day the order was placed (0-23).

For example, the first few rows show different items like "Bulgarian Yogurt" and "Organic Whole String Cheese" belonging to different orders.










